---
title: "CYO Cryptocurrency Report"
author: "Donal Medina"
date: "August 23, 2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
For this project we will predict the close price of cryptocurrency historical prices data set, we will develope a k-nearest neighbors (KNN) and random forest model and compare the perferomance.

# Analysis

## Install and load Packages 
```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
```

## Data Loading 
Create crypto_hist set. 
Cryptocurrency Historical Prices can found: https://www.kaggle.com/sudalairajkumar/cryptocurrencypricehistory
The Dataset can be download on my GitHub repository : https://github.com/dmedina-lab/cryptocurrency/raw/main/cryptocurrency_price.zip


```{r}
# Note: this process could take a couple of minutes
dlf <- tempfile()
download.file("https://github.com/dmedina-lab/cryptocurrency/raw/main/cryptocurrency_price.zip", dlf)

unzip(dlf, exdir = "crypto-price-hist")

crypto_hist <-
  list.files(path = "crypto-price-hist", pattern = "*.csv", full.names = T) %>% 
  map_df(~read_csv(.))

rm(dlf)
```

## Data Exploration
First we check for any NA value.  
```{r}
anyNA(crypto_hist)
```
General overview of dataset:   
```{r}
str(crypto_hist)
head(crypto_hist) %>% knitr::kable()
```
After loading the data set we start by looking at the data structure and type we can see that there is 37,082 observations and 10 variables.   
SNo: Incremental row number for each coin  
Name: Name of cryptcurrency  
Date : date of observation  
Open : Opening price on the given day  
High : Highest price on the given day  
Low : Lowest price on the given day   
Close : Closing price on the given day  
Volume : Volume of transactions on the given day  
Market Cap : Market capitalization in USD  

```{r}
summary(crypto_hist) %>% knitr::kable()
```
We can see the date range date from 2013-04-29 to 2021-07-06, the max Close price it is 63,503.46 usd and the average 987.12 usd.  
```{r}
crypto_hist %>% group_by(Name) %>% 
  summarize(count = n(), date_from= min(Date), date_to= max(Date) ) %>%
  arrange(desc(count)) %>% knitr::kable()
```
We can see there are 23 coins. Bitcoin, Litecoin, XRP, Dogecoin are the most old, since 2013. 

Top 5 newest cryptocurrency
```{r}
crypto_hist %>% group_by(Name) %>% 
  summarize(date_from= min(Date), date_to= max(Date) ) %>%
  arrange(desc(date_from)) %>%
  head(5) %>% knitr::kable()
```

Aave, Uniswap, Polkadot, Solana adn Cosmos are the newest coin, since 2019.  

Transformation of date formats to short date, year, month, day and price market relationship
```{r}
crypto_hist_p <- crypto_hist %>% 
  mutate(year = as.integer(format(Date, format="%Y"))
        , month = as.integer(format(Date, format="%m"))
        , day = as.integer(format(Date, format="%d"))
        , date_short = as.Date(format(Date, format="%Y-%m-%d"))
        , Price_Market = as.numeric(Close / Marketcap)
        )
```
Trend historical Close price for each Coin Name.   
```{r, echo=FALSE}
crypto_hist_p %>% ggplot(aes(date_short, Close, col=Name)) +
  geom_line() 
```
Boxplot historical Close price for each Coin Name.   
```{r, echo=FALSE}
crypto_hist_p %>% ggplot(aes(Name, Marketcap, fill=Name)) +
  geom_boxplot() 
```
Trend hisotorical Market Capitalization.   
```{r, echo=FALSE}
crypto_hist_p %>%  group_by(date_short) %>%
  summarize(mt = sum(Marketcap), dt = min(date_short)) %>%
  ggplot(aes(dt, mt)) +
  geom_line() 
```

Correlation between Close price and Market Capitalization.   
```{r}
crypto_hist_p %>% summarize(r = cor(Close, Marketcap)) %>% pull(r)
```
It is about 68.8%.  

Correlation between Close price and Volume.   
```{r}
crypto_hist_p %>% summarize(r = cor(Close, Volume)) %>% pull(r)
```
It is very low, about 28.5%.  

Now we will calculate the volatility of the cryptocurrencies en our dataset.
Volatility is best defined as the 30-day standard deviation of daily log returns annualized. 
With this in mind, high volatility is at 100% or above, medium volatility is between 50% and 100%, while low volatility is below 50%. 
Top 10 least Volatility (Anual).  
```{r}
crypto_hist_p %>% group_by(Name) %>% 
  summarize(avg_close = mean(Close),  sd_close = sd(Close), volatility = sqrt(365) * sd(diff(log(Close))) * 100) %>%
  arrange(volatility) %>%
  head(10) %>% knitr::kable()
```
Top 10 Most Volatility (Anual).  
```{r}
crypto_hist_p %>% group_by(Name) %>% 
  summarize(avg_close = mean(Close),  sd_close = sd(Close), volatility = sqrt(365) * sd(diff(log(Close))) * 100) %>%
  arrange(desc(volatility))%>%
  top_n(10)
```
We observe that Bitcoin es the third least volatility coin, The first and second least volatility are USD COin and Theter respectible, Both based on dollar.   

we will use Bitcoin in ours models.   
We will show the Volatility for year.
```{r}
crypto_hist_p %>% filter(Name=="Bitcoin") %>%group_by(year) %>% 
  summarize(avg_close = mean(Close),  sd_close = sd(Close), volatility = sqrt(365) * sd(diff(log(Close))) * 100) %>%
  arrange(volatility) %>% knitr::kable()
```
Trend of Bitcoin.
```{r, echo=FALSE}
crypto_hist_p %>% filter(Name == "Bitcoin") %>%
  ggplot(aes(date_short, Close)) +
  geom_line()  
```

## Data Wrangling

We will create a subset filtered by Bitcoin.
```{r}
btc <- crypto_hist_p %>% filter(Name == "Bitcoin")
```

Set seed.
```{r}
set.seed(1, sample.kind="Rounding")
```
Split the btc data set to train set(80%) and test set(20%).
```{r}
test_index_btc <- createDataPartition(y = btc$Close, times = 1, p = 0.2, list = FALSE)
train_set_btc <- btc[-test_index_btc,]
test_set_btc <- btc[test_index_btc,]
```


# Results

## MAPE (Mean Absolute Percentage Error)
```{r}
MAPE <- function(actual_price,pred_price){
  mape <- mean(abs((actual_price - pred_price)/actual_price))*100
  return (mape)
}
```
## RMSE (residual mean squared error) calculation funtion
```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2, na.rm = TRUE))
}
```

## First model : Random Forest
Fit the model.   
```{r}
set.seed(1, sample.kind="Rounding")
library(randomForest)

fit_rf <- randomForest(Close ~ Marketcap + Volume + Open + Low + High + date_short  + year + month + day, data=train_set_btc)

pred_rf <- predict(fit_rf, test_set_btc)
```

Evaluate the model with MAPE.  
```{r}
MAPE(test_set_btc$Close, pred_rf) 
```
Now with RMSE
```{r}
RMSE(test_set_btc$Close, pred_rf)
```

Plot the importance variable
```{r, echo=FALSE}
varImpPlot(fit_rf)  
```
The most important variables are Marketcap, Low, High, date_short and Open.

To avoid overtraing we will use only the most important variables to build the model.  
```{r}
rf_revised <- randomForest(Close ~ Marketcap + Low + High + date_short  + Open, data = train_set_btc)
print(rf_revised)
pred_rf_r <- predict(rf_revised, test_set_btc)
```
aguain evaluate the model with MAPE.  
```{r}
MAPE(test_set_btc$Close, pred_rf_r) 
```
Now with RMSE
```{r}
RMSE(test_set_btc$Close, pred_rf_r)
```

Plot the test results:
```{r, echo=FALSE}
test_set_btc %>%
  mutate(Close_hat = pred_rf_r ) %>%
  ggplot() +
  geom_point(aes(date_short, Close)) +
  geom_line(aes(date_short, Close_hat), col="red")  
```

## Second model : k-nearest neighbors (KNN)
Fit the model.   
```{r}
set.seed(1, sample.kind="Rounding")
train_knn <- train(Close ~ Marketcap + Low + High + date_short  + Open, method = "knn",
                   data = train_set_btc,
                   tuneGrid = data.frame(k = seq(1, 100, 2)))
ped_knn <- predict(train_knn, test_set_btc)
```

Evaluate the model with MAPE.  
```{r}
MAPE(test_set_btc$Close, ped_knn) 
```
Now with RMSE
```{r}
RMSE(test_set_btc$Close, ped_knn)
```
Plot the test results:
```{r, echo=FALSE}
test_set_btc %>%
  mutate(Close_hat = ped_knn ) %>%
  ggplot() +
  geom_point(aes(date_short, Close)) +
  geom_line(aes(date_short, Close_hat), col="red") 
```

# Conclusion
I have developed a k-nearest neighbors (KNN) and random forest model, the best MAPE given by the random forest model.  
The variability of the criptocurruncy its mostly high, in a future we can consider tunning this algorithms and use deep learning techniques like ARIMA, LSTM, etc. 
